---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r}
library (plyr)
```

## Loading and preprocessing the data

Assignment data set is included in the Git repository, forked from https://github.com/rdpeng/RepData_PeerAssessment1, in the file activity.zip. In my repository I've manually unpacked that to activity.csv. Load it:

```{r}
acts <- read.csv ('activity.csv')
```

The result is usable as is.

## What is mean total number of steps taken per day?

Plyr is a convenient tool to add up the steps taken in each day.

```{r}
spd <- ddply (acts,
              .(date), 
              summarize, 
              stepSum=sum (steps, na.rm=TRUE))
```

Display the result.

```{r}
hist (spd$stepSum, breaks=13)
```

The subject took a mean of `r mean (spd$stepSum)` steps and median of `r median (spd$stepSum)` steps per day.

## What is the average daily activity pattern?

Once again, Plyr is our friend.

```{r}
spi <- ddply (acts,
              .(interval), 
              summarize, 
              stepMean = mean(steps, na.rm=TRUE))
```

Display it:

```{r}
plot (spi$interval, spi$stepMean, type='l')
```

The busiest interval of the subject's day, on average was
`r spi$interval[which.max (spi$stepMean)]`.

## Imputing missing values

```{r}
which <- is.na (acts$steps)
```

The data set comprises many missing step count values: a total of
`r sum (which)`, or
`r 100.0 * sum (which) / length (which)`
percent of the total.

We use a two-stage strategy for imputing missing values to the data. First, we replace an NA at interval `i` with the mean number of steps in that interval over the rest of the data set. Unfortunately, many intervals are NA over the entire series; for those we impute the grand mean over all non-NA data.

```{r}
fixie <- acts
fixie$steps[which] <-
  spi$stepMean[spi$interval == fixie$interval[which]]
fixie$steps[is.na (fixie$steps)] <- 
  mean (acts$steps, na.rm=TRUE)
```

Fixed! Did it change anything? Here's the new histogram:
```{r}
fixieSPD <- ddply (fixie,
                   .(date), 
                   summarize, 
                   stepSum=sum (steps))
hist (fixieSPD$stepSum,
      breaks=13)
```

The outliers at the low end have gone away, which makes intuitive sense. Summary statistics also show this change in steps per day:

Data set  | Mean | Median
-- | -- | --
Raw only | `r mean (spd$stepSum, na.rm=TRUE)` | `r median (spd$stepSum, na.rm=TRUE)`
With imputed | `r mean (fixieSPD$stepSum)` | `r median (fixieSPD$stepSum)`


## Are there differences in activity patterns between weekdays and weekends?

First, let's distinguish weekdays from weekends.

```{r}
wd <- weekdays (as.Date (fixie$date))
fixie$dayTypes <- factor (wd,
                          levels <- c ('weekday', 'weekend'))
fixie$dayTypes[] <- 'weekday'
fixie$dayTypes[wd %in% c ('Saturday', 'Sunday')] <- 'weekend'
```

Now, compute summaries and plot the two:

```{r}
library (lattice)
spi <- ddply (fixie,
              .(interval, dayTypes), 
              summarize, 
              stepMean = mean(steps, na.rm=TRUE))
xyplot (stepMean ~ interval | dayTypes, 
        data=spi,
        type='l',
        layout=c (1, 2))
```

Conclusion: the subjects slept later on weekends. Good on 'em!
